Pipit on the Post: Proving Pre- and Post-conditions of Reactive Systems

### 2024-01-16 2024pipit/section/30-Core.tex
diagram

@@

% For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table \autoref{s:core:primitive-tables}.

@@

\subsection{Dynamic semantics}
\input{figures/core-bigstep.tex}

@@

Contracts involve two proofs: one for the definition and one for the instantiation.
To prove that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid, we show that for all streaming histories $\Sigma$, assuming the rely is always true under the history ($\bigstepalways{\Sigma}{\erely}$), then the body always satisfies the guarantee ($\bigstepalways{\Sigma}{\eguar[x := \ebody]}$).
Additionally, we can also assume that the valid properties in all three components hold, and we must also show that the unknown properties are valid.
The fact that the checked semantics refers to a particular $\Sigma$ is significant here: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.
\TODO{WAFFLE? This is expanded on in \autoref{s:core:blessing}; kill here?}

>>>>

\input{figures/core-structure.tex}

\autoref{f:core:structure} shows the architecture of Pipit.
On the left-hand-side, the surface syntax embedded in \fstar{} is shown; this includes some Pipit-specific syntactic sugar.
The translation from the surface syntax to the core language is trusted.
There are two targets from the core language: abstract transition systems for verification, and executable transition systems for extraction to C.
The translation from the abstract systems is verified to be an abstraction according to the dynamic semantics (\autoref{s:core:dynamic}).
One property that still remains to be proven is that if the abstract transition system says a contract or checked property holds, then the same contract or property holds on the original semantics; this is denoted as negated entailment in the figure ($\not\vdash$).
The translation to executable transition system is proven to be semantics preserving, as is the subsequent translation to \lowstar{}.
The translation from \lowstar{} to C is external to this paper and forms part of our trusted computing base.

@@

For the presentation of the formal grammar here, we consider only a fixed set of values and primitives; in practice, the implementation is parameterised by a primitive table which we extend with immutable array operations for the TTCAN driver logic in \autoref{s:evaluation}.

@@

\subsection{Dynamic semantics}
\label{s:core:dynamic}
\input{figures/core-bigstep.tex}

@@

Contracts involve two proofs: one for the definition and one for the instantiation.
To prove that a contract definition $\xcontractP{\PStatus}{\erely}{\ebody}{\rawbind{x}\eguar[x]}$ is valid, we show that for all streaming histories $\Sigma$, assuming the rely is always true under the history ($\bigstepalways{\Sigma}{\erely}$), then the body always satisfies the guarantee ($\bigstepalways{\Sigma}{\eguar[x := \ebody]}$).
Additionally, we can also assume that the valid properties in all three components hold, and we must also show that the unknown properties are valid.
The fact that the checked semantics refers to a particular $\Sigma$ is significant here: it allows the proof of contract validity to only consider streaming histories where the rely actually holds.

### 2024-01-16 2024pipit/section/60-Evaluation.tex
diagram

As a preliminary evaluation of Pipit, we have implemented the high-level logic of a time-triggered Controller Area Network (CAN) bus driver~\cite{ISO11898_4}.
The CAN bus is commonly found in safety-critical automotive and industrial settings.
The time-triggered network architecture defines a static schedule of network traffic.
All nodes on the network must adhere to the same schedule, which significantly increases the reliability of periodic messages~\cite{fuehrer2001time}.

@@

\begin{figure}
  \center
\begin{tabular}{r|rr|rr|rr}
  & \multicolumn{4}{c|}{Kind2} & Pipit \\
  & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
  size & wall-clock & CPU & wall-clock & CPU & wall-clock & CPU \\
  \hline
  1 & 2s &  3s  & 6s & 12s & 6s & 6s \\
  2 & 3s &  3s  & 8s & 12s & 6s & 6s \\
  4 & 5s & 11s  & 12s & 24s & 6s & 6s \\
  8 & 8s & 13s  & 82s & 90s & 6s & 6s \\
  16 & 125s & 276s & \multicolumn{2}{c|}{error} & 6s & 6s \\
  32 & \multicolumn{2}{c|}{error} & \multicolumn{2}{c|}{error} & 6s & 6s \\
\end{tabular}
\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification.}
\label{f:evaluation:kind2-runtime}
\end{figure}

>>>>

To evaluate Pipit, we have implemented the high-level logic of a time-triggered Controller Area Network (CAN) bus driver~\cite{ISO11898_4}.
The CAN bus is commonly found in safety-critical automotive and industrial settings.
The time-triggered network architecture defines a static schedule of network traffic.
All nodes on the network must adhere to the same schedule, which significantly increases the reliability of periodic messages~\cite{fuehrer2001time}.

@@

\begin{figure}
  \center
\begin{tabular}{r|rr|rr|rr}
  & \multicolumn{4}{c|}{Kind2} & Pipit \\
  & \multicolumn{2}{c|}{simple enable-set} & \multicolumn{2}{c|}{full enable-set} & \\
  size & wall-clock & CPU & wall-clock & CPU & wall-clock & CPU \\
  \hline
  1 & 2s &  3s  & 6s & 12s & 6s & 6s \\
  2 & 3s &  3s  & 8s & 12s & 6s & 6s \\
  4 & 5s & 11s  & 12s & 24s & 6s & 6s \\
  8 & 8s & 13s  & 82s & 90s & 6s & 6s \\
  16 & 125s & 276s & \multicolumn{2}{c|}{error} & 6s & 6s \\
  32 & \multicolumn{2}{c|}{error} & \multicolumn{2}{c|}{error} & 6s & 6s \\
\end{tabular}
\caption{Verification time for trigger-fetch; simple enable-set uses a simplified version of the enable-set, while full enable-set uses bitwise arithmetic as in the TTCAN specification. The verification time for Pipit is a once-and-for-all proof that is parametric in the size of the array.}
\label{f:evaluation:kind2-runtime}
\end{figure}

### 2024-01-16 2024pipit/section/80-Conclusion.tex
diagram

We have presented Pipit, a verified compiler and proof system for reactive systems.
Our implementation of the TTCAN driver logic shows that Pipit can express programs which are currently unsupported by other verified Lustre compilers.
Pipit can also verify high-level program properties which are difficult to express and prove in existing Lustre model-checkers.
Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
We also introduced a checked semantics, which describes the proof obligations of a program; in future work, we intend to verify that any translated property the user proves on the transition system holds for the original property on the checked semantics.

In the future, we intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
Clocks are important for composing complex systems together and avoiding unnecessary computation.

>>>>

We have presented Pipit, a verified compiler and proof system for reactive systems.
Our implementation of the TTCAN driver logic shows that, by embedding pure \fstar{} functions for array operations, Pipit can express programs which are currently unsupported by other verified Lustre compilers.
Pipit can also verify high-level program properties which are difficult to express and prove in existing Lustre model-checkers.
Our development includes verified translations to both abstract and executable transition systems; both are shown to preserve the dynamic semantics.
We also introduced a checked semantics, which describes the proof obligations of a program; in future work, we intend to verify that the proof obligations generated by the abstract transition system match the checked semantics.

In the future, we intend to verify the remainder of the TTCAN driver logic.
We also intend to increase the expressivity of Pipit by adding \emph{clocks}, which are used to describe partially-defined streams~\cite{caspi1995functional}.
Clocks are important for composing complex systems together and avoiding unnecessary computation; clocks may be useful if it becomes necessary to optimise the runtime of the TTCAN driver.
